\tit

### 2021-02-11 paper.tex
rebuilding paper and talk with shared Agda sources

\usepackage[links]{agda}% references

@@

\renewcommand{\AgdaFontStyle}[1]{\text{#1}}

@@

%% \nc\tit{Working Title}
\nc\tit{Symbolic and Automatic Differentiation of Languages}

@@

A simple type isomorphism property captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level languages (and indeed functions from lists to any type).
These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing almost exactly the same code but in dual arrangements.
The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.
The regular expression version corresponds to symbolic differentiation, while the trie version corresponds to automatic differentiation.

@@

%% \input{NewIntroduction}
\input{Introduction}
%% \input{Generation}
\input{Predicate}
%% \input{AltDec}
%% \input{AltCalc}
\input{Predicate/Algebra}
\input{Nullability}
\input{Decidability}
\input{Calculus}
% \input{NonTrie}
\input{Regex}
\input{Trie}
\input{ProvingProperties}

%% \input{BeyondPredicates}
%% \input{Predicate/Examples}
%% \input{Relation}

\workingHere
\note{To do: relate regular expressions and tries to symbolic and automatic differentiation; reflections, related work, future work.}

%% \definecolor{mylinkcolor}{rgb}{0,0.4,0}
%% \hypersetup{
%%   linkcolor  = mylinkcolor,
%%   citecolor  = mylinkcolor,
%%   urlcolor   = mylinkcolor,
%%   colorlinks = true
%% }

>>>>

\usepackage{agda}% references

@@

%% \renewcommand{\AgdaFontStyle}[1]{\text{#1}}

@@

%% \nc\tit{Working Title}
\nc\tit{%Functional Pearl:
Symbolic and Automatic Differentiation of Languages}

@@

A simple type isomorphism property captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level languages (and indeed functions from lists to any type).
These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.
The regular expression version corresponds to symbolic differentiation, while the trie version corresponds to automatic differentiation.

@@

\sectionl{Specifying Languages}

Languages are usually formalized either set-theoretically as a set of strings \needcite{} or operationally as a parser \needcite{}.
Alternatively, we can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters''):
\Language{Lang}
\figrefdef{Lang-ops}{Language operations}{\Language{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).

\note{Explain the types used here and their logical interpretation under the Curry-Howard isomorphism.}

Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsing}.
(If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
For instance, a proof of {\AB{w} ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen.
Likewise, a proof of {w ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} ∈ \AB{P}} and {\AB{v} ∈ \AB{P}}.

The use of type-level predicates makes for simple, direct specification, including the use of existential quantification.
As described in \secref{Language Algebra}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
These definitions do not, however, give us decidable parsing, because general type inhabitation amounts to theorem proving.

\sectionl{Decomposing Languages}

In order to convert languages into parsers, it will help to decompose languages---and indeed any function from lists---into a regular form.
Since lists are defined by an algebraic data type, we can decompose languages according to constructor.
